---
title: "21-deseq2-2021"
output: html_document
date: "2024-10-11"
---
Rmd to take gene count matrix from [code/20-hisat2-genecount-matrices.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/20-hisat2-genecount-matrices.Rmd) through `DESeq2` to get a list or lists of differentially expressed genes. 

This was done in a previous Rmd ([code/18-deseq2_2021_2022.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/18-deseq2_2021_2022.Rmd)), but doing in a new document because I'm using a new count matrix. 

```{r}
library(DESeq2)
library(dplyr)
library(ggplot2)
library(pheatmap)
```


| library_ID | star_ID | treatment         | sample_date | experiment_day |
|------------|---------|-------------------|-------------|----------------|
| PSC.56     | E06     | control (10/5/21) | 10/14/21    | 9              |
| PSC.52     | E11     | control (10/5/21) | 10/14/21    | 9              |
| PSC.54     | E18     | control (10/5/21) | 10/14/21    | 9              |
| PSC.61     | E09     | control (10/5/21) | 10/15/21    | 10             |
| PSC.64     | H05     | control (10/5/21) | 10/15/21    | 10             |
| PSC.73     | H03     | control (10/5/21) | 10/16/21    | 11             |
| PSC.76     | E10     | control (10/5/21) | 10/17/21    | 12             |
| PSC.81     | E07     | control (10/5/21) | 10/18/21    | 13             |
| PSC.59     | H01     | exposed (10/5/21) | 10/14/21    | 9              |
| PSC.57     | H06     | exposed (10/5/21) | 10/14/21    | 9              |
| PSC.69     | H04     | exposed (10/5/21) | 10/15/21    | 10             |
| PSC.67     | H09     | exposed (10/5/21) | 10/15/21    | 10             |
| PSC.71     | H18     | exposed (10/5/21) | 10/15/21    | 10             |
| PSC.75     | H08     | exposed (10/5/21) | 10/16/21    | 11             |
| PSC.78     | H10     | exposed (10/5/21) | 10/17/21    | 12             |
| PSC.83     | H07     | exposed (10/5/21) | 10/18/21    | 13             |

Read in gene count matrix:
```{r}
counts21 <- read.csv("../data/gene_count_matrix_2021.csv")
head(counts21)
```
calling subset of libraries "exp" short for "experiment" because these libraries are part of the experiment of interest. 
```{r}
exp <- select(counts21, "gene_id", "PSC.56", "PSC.52", "PSC.54", "PSC.61", "PSC.64", "PSC.73", "PSC.76", "PSC.81", "PSC.59", "PSC.57", "PSC.69", "PSC.67", "PSC.71", "PSC.75", "PSC.78", "PSC.83")
head(exp)
```
23,464 rows, 17 columns (16 libraries)

set column one (gene_id) as rownames: 
```{r}
exp <- exp %>% remove_rownames %>% column_to_rownames(var='gene_id')
```



Make a data frame for the comparison: 
```{r}
colData <- data.frame(condition=factor(c("control", "control", "control", "control", "control", "control", "control", "control", "exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "exposed")),
                      type=factor(rep("paired-end",16)))
rownames(colData) <- colnames(exp)
dds <- DESeqDataSetFromMatrix(countData = exp,
                              colData = colData,
                              design = ~ condition)
```


```{r}
# Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colData) %in% colnames(exp)) #should return TRUE
```

```{r}
exp <- exp[, rownames(colData)]
all(rownames(colData) == colnames(exp)) # This should also return TRUE
```

```{r}
# Create a DESeqDataSet from count matrix and labels
dds <- DESeqDataSetFromMatrix(countData = exp,
                              colData = colData, design = ~ condition)


# Run the default analysis for DESeq2 and generate results table
dds <- DESeq(dds)
deseq2.res <- results(dds)

# Sort by adjusted p-value and display
resOrdered <- deseq2.res[order(deseq2.res$padj), ]
vsd <- vst(dds, blind = FALSE)
expPCA <- plotPCA(vsd, intgroup = "condition")
nudge <- position_nudge(y = 4)
expPCA + geom_text(aes(label = name), position = nudge)
```
```{r}
# Select top 50 differentially expressed genes
res <- results(dds)
res_ordered <- res[order(res$padj), ]
top_genes <- row.names(res_ordered)[1:50]

# Extract counts and normalize
counts <- counts(dds, normalized = TRUE)
counts_top <- counts[top_genes, ]

# Log-transform counts
log_counts_top <- log2(counts_top + 1)

# Generate heatmap
pheatmap(log_counts_top, scale = "row")

```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ])
```
6938 degs

```{r}
expg <- deseq2.res
# The main plot
plot(expg$baseMean, expg$log2FoldChange, pch=20, cex=0.45, ylim=c(-3, 3), log="x", col="darkgray",
     main="DEG SSWD Exposed  (pval <= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
expg.sig <- deseq2.res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ]
points(expg.sig$baseMean, expg.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```





