---
title: "22-deseq2-2022"
output: html_document
date: "2024-10-11"
---
Rmd to take gene count matrix from [code/20-hisat2-genecount-matrices.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/20-hisat2-genecount-matrices.Rmd) through `DESeq2` to get a list or lists of differentially expressed genes. 

This was done in a previous Rmd ([code/15-hisat2-deseq2-summer2022.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/15-hisat2-deseq2-summer2022.Rmd)), but doing in a new document because I'm using a new count matrix. 

```{r}
library(DESeq2)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(tidyverse)
```


| library_ID | star_ID | star_age_class | treatment | experiment_day | total_armdrop | total_armtwist | disease_sign  |
|------------|---------|----------------|-----------|----------------|---------------|----------------|---------------|
| PSC.0186   |      15 | adult          | control   |             11 |             0 |              0 | healthy       |
| PSC.0203   |      31 | adult          | control   |             12 |             0 |              0 | healthy       |
| PSC.0209   |       9 | adult          | control   |             13 |             0 |              0 | healthy       |
| PSC.0177   |      34 | juvenile       | control   |             11 |             0 |              0 | healthy       |
| PSC.0219   |      63 | juvenile       | control   |             14 |             0 |              0 | healthy       |
| PSC.0230   |      65 | juvenile       | control   |             15 |             0 |              0 | healthy       |
| PSC.0174   |      30 | adult          | exposed   |             10 |             1 |              6 | first_armdrop |
| PSC.0190   |      10 | adult          | exposed   |             11 |             1 |              0 | first_armdrop |
| PSC.0231   |      13 | adult          | exposed   |             15 |             1 |              5 | first_armdrop |
| PSC.0187   |      57 | juvenile       | exposed   |             11 |             1 |              1 | first_armdrop |
| PSC.0188   |      38 | juvenile       | exposed   |             11 |             1 |              0 | first_armdrop |
| PSC.0228   |      61 | juvenile       | exposed   |             14 |             1 |              0 | first_armdrop |

Read in gene count matrix:
```{r}
counts22 <- read.csv("../data/gene_count_matrix_2022.csv")
head(counts22)
```

subset just the counts for the libraries specified in the table above:
```{r}
counts22sub <- select(counts22, gene_id, PSC.0228, PSC.0187, PSC.0188, PSC.0174, PSC.0190, PSC.0231, PSC.0230, PSC.0219, PSC.0177, PSC.0186, PSC.0209, PSC.0203)
head(counts22sub)
```
23464 rows, 13 columns (12 libraries) 

set column one (gene_id) as rownames: 
```{r}
counts22sub <- counts22sub %>% remove_rownames %>% column_to_rownames(var='gene_id')
head(counts22sub)
```
## Compare Exposed to Control 

Start out by just comparing exposed and control without doing anything with age. 
```{r}
colData <- data.frame(condition=factor(c("exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "control", "control", "control", "control", "control", "control")),
                      type=factor(rep("paired-end",12)))
rownames(colData) <- colnames(counts22sub)
dds <- DESeqDataSetFromMatrix(countData = counts22sub,
                              colData = colData,
                              design = ~ condition)
```


```{r}
# Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colData) %in% colnames(counts22sub)) #should return TRUE
```

```{r}
counts22sub <- counts22sub[, rownames(colData)]
all(rownames(colData) == colnames(counts22sub)) # This should also return TRUE
```

```{r}
# Create a DESeqDataSet from count matrix and labels
dds <- DESeqDataSetFromMatrix(countData = counts22sub,
                              colData = colData, design = ~ condition)


# Run the default analysis for DESeq2 and generate results table
dds <- DESeq(dds)
deseq2.res <- results(dds)

# Sort by adjusted p-value and display
resOrdered <- deseq2.res[order(deseq2.res$padj), ]
vsd <- vst(dds, blind = FALSE)
exp22PCA <- plotPCA(vsd, intgroup = "condition")
nudge <- position_nudge(y = 4)
exp22PCA + geom_text(aes(label = name), position = nudge)
```

```{r}
# Select top 50 differentially expressed genes
res <- results(dds)
res_ordered <- res[order(res$padj), ]
top_genes <- row.names(res_ordered)[1:50]

# Extract counts and normalize
counts <- counts(dds, normalized = TRUE)
counts_top <- counts[top_genes, ]

# Log-transform counts
log_counts_top <- log2(counts_top + 1)

# Generate heatmap
pheatmap(log_counts_top, scale = "row")

```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ])
```
6237 degs!
just comparing exposed and control

```{r}
exp22 <- deseq2.res
# The main plot
plot(exp22$baseMean, exp22$log2FoldChange, pch=20, cex=0.45, ylim=c(-3, 3), log="x", col="darkgray",
     main="DEG SSWD Exposed  (pval <= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
exp22.sig <- deseq2.res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ]
points(exp22.sig$baseMean, exp22.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

Write out differentially expressed gene list: 
```{r}
#write.table(exp22.sig, "../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed.tab", sep = '\t', row.names = T, quote = F)
```
comment out the code - wrote out 2024-10-16

## Compare Exposed to Control Taking Age Into Account

```{r}
colDataCvEwA <- data.frame(condition=factor(c("exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "control", "control", "control", "control", "control", "control")), 
                             type=factor(rep("paired-end", 12)),
                             age=factor(c("juvenile", "juvenile", "juvenile", "adult", "adult", "adult", "juvenile", "juvenile", "juvenile", "adult", "adult", "adult")))
rownames(colDataCvEwA) <- colnames(counts22sub)
```

```{r}
# Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colDataCvEwA) %in% colnames(counts22sub)) #should return TRUE
```

```{r}
counts22sub <- counts22sub[, rownames(colDataCvEwA)]
all(rownames(colDataCvEwA) == colnames(counts22sub)) # This should also return TRUE
```

modelled after crab paper https://github.com/RobertsLab/paper-tanner-crab/blob/master/scripts/DESeq.Rmd 

```{r}
# Create a DESeqDataSet from count matrix and labels, with the design taking age into account
ddsCvEwA <- DESeqDataSetFromMatrix(countData = counts22sub,
                                     colData = colDataCvEwA, 
                                     design = ~ condition + age)
```

Check levels of age and condition 
```{r}
levels(ddsCvEwA$age)
```


```{r}
levels(ddsCvEwA$condition)
```

note: `DESeq2` automatically puts the levels in alphabetical order and the first listed level is the reference level for the factor. 

So, Control is reference. 

The following will pull the results from `condition` because that is our variable of interest. This tells us how age contributes to the DEGs
```{r}
design(ddsCvEwA) <- formula(~ age + condition)
ddsCvEwA <- DESeq(ddsCvEwA)
```

Access results:
```{r}
ddsCvEwA.res <- results(ddsCvEwA)
head(ddsCvEwA.res)
```


```{r}
summary(ddsCvEwA.res)
```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(ddsCvEwA.res[!is.na(ddsCvEwA.res$padj) & ddsCvEwA.res$padj <= 0.05, ])
```
6202

```{r}
CvEwA <- ddsCvEwA.res
# The main plot
plot(CvEwA$baseMean, CvEwA$log2FoldChange, pch=20, cex=0.45, ylim=c(-15, 15), log="x", col="darkgray",
     #main="Infection Status  (pval </= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
CvEwA.sig <- ddsCvEwA.res[!is.na(ddsCvEwA.res$padj) & ddsCvEwA.res$padj <= 0.05, ]
points(CvEwA.sig$baseMean, CvEwA.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

Write out differentially expressed gene list: 
```{r}
#write.table(CvEwA.sig, "../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed_withAge.tab", sep = '\t', row.names = T, quote = FALSE)
```

comment out 2024-10-16

## Age Contrast:

Now need to perform a contrast to see if there is a difference between these groups as it relates to age. 

In the multifactor section of the `DESeq2` manual:                
The contrast argument of the function _results_ needs a character vector of three componenets: the name of the variable (in this case "age"), and the name of the factor level for the numerator of the log2 ratio (juvenile) and the denominator (adult) 

A **contrast** is a linear combination of estimated log2 fold changes. Can be used to test if differences between groups are equal to zero. ‹        
```{r}
resultsNames(ddsCvEwA)
```

```{r}
deseq2.resage <- results(ddsCvEwA,
                          contrast = c("age", "juvenile",  "adult"))
head(deseq2.resage)
```

```{r}
age <- deseq2.resage
# The main plot
plot(age$baseMean, age$log2FoldChange, pch=20, cex=0.45, ylim=c(-15, 15), log="x", col="darkgray",
     #main="Age  (pval </= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
age.sig <- deseq2.resage[!is.na(deseq2.resage$padj) & deseq2.resage$padj <= 0.05, ]
points(age.sig$baseMean, age.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(deseq2.resage[!is.na(deseq2.resage$padj) & deseq2.resage$padj <= 0.05, ])
```

82

```{r}
summary(deseq2.resage)
```

write out DEGs from age contrast:
```{r}
#write.table(age.sig, "../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed_ageContrast.tab", sep = '\t', row.names = T, quote = FALSE)
```

wrote out 2024-10-16

## Compare Exposed to Control Taking Size Into Account (as a proxy for age)
2025-06-03
NOTE --> DOES NOT WORK - commenting out the code chunks

#```{r}
colDataCvEwS <- data.frame(condition=factor(c("exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "control", "control", "control", "control", "control", "control")), 
                             type=factor(rep("paired-end", 12)),
                             size=factor(c("10cm", "10cm", "10cm", "32cm", "44cm", "34cm", "12cm", "8cm", "14cm", "36cm", "38cm", "42cm")))
rownames(colDataCvEwS) <- colnames(counts22sub)
#```


#```{r}
# Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colDataCvEwS) %in% colnames(counts22sub)) #should return TRUE
#```

#```{r}
counts22sub <- counts22sub[, rownames(colDataCvEwS)]
all(rownames(colDataCvEwS) == colnames(counts22sub)) # This should also return TRUE
#```

modelled after crab paper https://github.com/RobertsLab/paper-tanner-crab/blob/master/scripts/DESeq.Rmd 

#```{r}
# Create a DESeqDataSet from count matrix and labels, with the design taking age into account
ddsCvEwS <- DESeqDataSetFromMatrix(countData = counts22sub,
                                     colData = colDataCvEwS, 
                                     design = ~ condition + size)
#```

Check levels of age and condition 
#```{r}
levels(ddsCvEwS$size)
#```


#```{r}
levels(ddsCvEwS$condition)
#```

note: `DESeq2` automatically puts the levels in alphabetical order and the first listed level is the reference level for the factor. 

So, Control is reference. 

The following will pull the results from `condition` because that is our variable of interest. This tells us how age contributes to the DEGs
#```{r}
design(ddsCvEwS) <- formula(~ size + condition)
ddsCvEwS <- DESeq(ddsCvEwS)
#```

Access results:
#```{r}
ddsCvEwS.res <- results(ddsCvEwS)
head(ddsCvEwS.res)
#```


#```{r}
summary(ddsCvEwS.res)
#```

#```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(ddsCvEwS.res[!is.na(ddsCvEwS.res$padj) & ddsCvEwS.res$padj <= 0.05, ])
#```
6202

#```{r}
CvEwS <- ddsCvEwS.res
# The main plot
plot(CvEwA$baseMean, CvEwS$log2FoldChange, pch=20, cex=0.45, ylim=c(-15, 15), log="x", col="darkgray",
     #main="Infection Status  (pval </= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
CvEwS.sig <- ddsCvEwS.res[!is.na(ddsCvEwS.res$padj) & ddsCvEwS.res$padj <= 0.05, ]
points(CvEwS.sig$baseMean, CvEwS.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
#```

#################################################################

# Summer 2025
Revisions from Reviewer Comments

Tutorial linked to by reviewer: https://www.atakanekiz.com/technical/a-guide-to-designs-and-contrasts-in-DESeq2/ 

Section 3 "Two factors with interaction" is the right section for my data. Their model is based on a treatment, and comparing it between males and females. So there's control and exposed females, and control and exposed males. In our case, we have control and exposed adults, and control and exposed juveniles. 

I'll model it off of what is in the tutorial. 

1. Read in the data: 
Read in gene count matrix:
```{r}
counts22 <- read.csv("../data/gene_count_matrix_2022.csv")
head(counts22)
```

subset just the counts for the libraries specified in the table above:
```{r}
counts22sub <- select(counts22, gene_id, PSC.0228, PSC.0187, PSC.0188, PSC.0174, PSC.0190, PSC.0231, PSC.0230, PSC.0219, PSC.0177, PSC.0186, PSC.0209, PSC.0203)
head(counts22sub)
```

set column one (gene_id) as rownames: 
```{r}
counts22sub <- counts22sub %>% remove_rownames %>% column_to_rownames(var='gene_id')
head(counts22sub)
```

Create the Deseq data set - set the conditions and the age 
CvEwA = Control vs Exposed with Age
```{r}
colDataCvEwA <- data.frame(condition=factor(c("exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "control", "control", "control", "control", "control", "control")), 
                             type=factor(rep("paired-end", 12)),
                             age=factor(c("juvenile", "juvenile", "juvenile", "adult", "adult", "adult", "juvenile", "juvenile", "juvenile", "adult", "adult", "adult")))
rownames(colDataCvEwA) <- colnames(counts22sub)
```

New addition to below code chunk from the tutorial: `design = ~ condition + age + condition:age`

```{r}
# Create a DESeqDataSet from count matrix and labels, with the design taking age into account
ddsCvEwA <- DESeqDataSetFromMatrix(countData = counts22sub,
                                     colData = colDataCvEwA, 
                                     design = ~ condition + age + condition:age)
```

```{r}
ddsCvEwA <- DESeq(ddsCvEwA)
resultsNames(ddsCvEwA)
```


Check levels of age and condition 
```{r}
levels(ddsCvEwA$age)
```


```{r}
levels(ddsCvEwA$condition)
```


2. Follow section 3.1 from the tutorial 

create a function called `contraster`:

```{r}
contraster <- function(dds,    # should contain colData and design
                       group1, # list of character vectors each with 2 or more items 
                       group2, # list of character vectors each with 2 or more items
                       weighted = F
){
  
  
  mod_mat <- model.matrix(design(dds), colData(dds))
  
  grp1_rows <- list()
  grp2_rows <- list()
  
  
  for(i in 1:length(group1)){
    
    grp1_rows[[i]] <- colData(dds)[[group1[[i]][1]]] %in% group1[[i]][2:length(group1[[i]])]
    
  }
  
  
  for(i in 1:length(group2)){
    
    grp2_rows[[i]] <- colData(dds)[[group2[[i]][1]]] %in% group2[[i]][2:length(group2[[i]])]
    
  }
  
  grp1_rows <- Reduce(function(x, y) x & y, grp1_rows)
  grp2_rows <- Reduce(function(x, y) x & y, grp2_rows)
  
  mod_mat1 <- mod_mat[grp1_rows, ,drop=F]
  mod_mat2 <- mod_mat[grp2_rows, ,drop=F]
  
  if(!weighted){
    
    mod_mat1 <- mod_mat1[!duplicated(mod_mat1),,drop=F]
    mod_mat2 <- mod_mat2[!duplicated(mod_mat2),,drop=F]
    
  }
  
  return(colMeans(mod_mat1)-colMeans(mod_mat2))
  
  
}
```


```{r}
# adult vs juvenile (in the control group) 
res1 <- results(ddsCvEwA, 
                contrast = contraster(ddsCvEwA,
                                      group1 = list(c("age", "adult"), c("condition", "control")),
                                      group2 = list(c("age", "juvenile", c("condition", "control")))))
head(res1,3)
```
```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res1[!is.na(res1$padj) & res1$padj <= 0.05, ])
```
982 degs 


```{r}
# adult vs juvenile (in exposed group)
res2 <- results(ddsCvEwA, 
                contrast = contraster(ddsCvEwA,
                                      group1 = list(c("age", "adult"), c("condition", "exposed")),
                                      group2 = list(c("age", "juvenile", c("condition", "exposed")))))
head(res2,3)
```
```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res2[!is.na(res2$padj) & res2$padj <= 0.05, ])
```
2797

```{r}
# exposed vs control (for adults)
res3 <- results(ddsCvEwA, 
                contrast = contraster(ddsCvEwA,
                                      group1 = list(c("age", "adult"), c("condition", "exposed")),
                                      group2 = list(c("age", "adult", c("condition", "control")))))
head(res3,3)
```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res3[!is.na(res3$padj) & res3$padj <= 0.05, ])
```
4841

```{r}
# exposed vs control (for juveniles)
res4 <- results(ddsCvEwA, 
                contrast = contraster(ddsCvEwA,
                                      group1 = list(c("age", "juvenile"), c("condition", "exposed")),
                                      group2 = list(c("age", "juvenile", c("condition", "control")))))
head(res4,3)
```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res4[!is.na(res4$padj) & res4$padj <= 0.05, ])
```
2927

## See interaction 
```{r}
# adult vs juvenile in exposed group 
diff1 <- contraster(ddsCvEwA,
                    group1 = list(c("age", "adult"), c("condition", "exposed")),
                    group2 = list(c("age", "juvenile"), c("condition", "exposed")))

# adult vs juvenile in control group
diff2 <- contraster(ddsCvEwA,
                    group1 = list(c("age", "adult"), c("condition", "control")),
                    group2 = list(c("age", "juvenile"), c("condition", "control")))

res5 <- results(ddsCvEwA,
                contrast = diff1-diff2)

head(res5, 3)
```


```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res5[!is.na(res5$padj) & res5$padj <= 0.05, ])
```
140 DEGs 

```{r}
plot5 <- res5
# The main plot
plot(plot5$baseMean, plot5$log2FoldChange, pch=20, cex=0.45, ylim=c(-3, 3), log="x", col="darkgray",
     main="DEGs of interaction between age and exposure (pval <= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
plot5.sig <- res5[!is.na(res5$padj) & res5$padj <= 0.05, ]
points(plot5.sig$baseMean, plot5.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

Write out differentially expressed gene list: 
```{r}
#write.table(plot5.sig, "../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed-with-age-interaction.tab", sep = '\t', row.names = T, quote = F)
```

