---
title: "22-deseq2-2022"
output: html_document
date: "2024-10-11"
---
Rmd to take gene count matrix from [code/20-hisat2-genecount-matrices.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/20-hisat2-genecount-matrices.Rmd) through `DESeq2` to get a list or lists of differentially expressed genes. 

This was done in a previous Rmd ([code/15-hisat2-deseq2-summer2022.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/15-hisat2-deseq2-summer2022.Rmd)), but doing in a new document because I'm using a new count matrix. 

```{r}
library(DESeq2)
library(dplyr)
library(ggplot2)
library(pheatmap)
```


| library_ID | star_ID | star_age_class | treatment | experiment_day | total_armdrop | total_armtwist | disease_sign  |
|------------|---------|----------------|-----------|----------------|---------------|----------------|---------------|
| PSC.0186   |      15 | adult          | control   |             11 |             0 |              0 | healthy       |
| PSC.0203   |      31 | adult          | control   |             12 |             0 |              0 | healthy       |
| PSC.0209   |       9 | adult          | control   |             13 |             0 |              0 | healthy       |
| PSC.0177   |      34 | juvenile       | control   |             11 |             0 |              0 | healthy       |
| PSC.0219   |      63 | juvenile       | control   |             14 |             0 |              0 | healthy       |
| PSC.0230   |      65 | juvenile       | control   |             15 |             0 |              0 | healthy       |
| PSC.0174   |      30 | adult          | exposed   |             10 |             1 |              6 | first_armdrop |
| PSC.0190   |      10 | adult          | exposed   |             11 |             1 |              0 | first_armdrop |
| PSC.0231   |      13 | adult          | exposed   |             15 |             1 |              5 | first_armdrop |
| PSC.0187   |      57 | juvenile       | exposed   |             11 |             1 |              1 | first_armdrop |
| PSC.0188   |      38 | juvenile       | exposed   |             11 |             1 |              0 | first_armdrop |
| PSC.0228   |      61 | juvenile       | exposed   |             14 |             1 |              0 | first_armdrop |

Read in gene count matrix:
```{r}
counts22 <- read.csv("../data/gene_count_matrix_2022.csv")
head(counts22)
```

subset just the counts for the libraries specified in the table above:
```{r}
counts22sub <- select(counts22, gene_id, PSC.0228, PSC.0187, PSC.0188, PSC.0174, PSC.0190, PSC.0231, PSC.0230, PSC.0219, PSC.0177, PSC.0186, PSC.0209, PSC.0203)
head(counts22sub)
```
23464 rows, 13 columns (12 libraries) 

set column one (gene_id) as rownames: 
```{r}
counts22sub <- counts22sub %>% remove_rownames %>% column_to_rownames(var='gene_id')
head(counts22sub)
```
## Compare Exposed to Control 

Start out by just comparing exposed and control without doing anything with age. 
```{r}
colData <- data.frame(condition=factor(c("exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "control", "control", "control", "control", "control", "control")),
                      type=factor(rep("paired-end",12)))
rownames(colData) <- colnames(counts22sub)
dds <- DESeqDataSetFromMatrix(countData = counts22sub,
                              colData = colData,
                              design = ~ condition)
```


```{r}
# Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colData) %in% colnames(counts22sub)) #should return TRUE
```

```{r}
counts22sub <- counts22sub[, rownames(colData)]
all(rownames(colData) == colnames(counts22sub)) # This should also return TRUE
```

```{r}
# Create a DESeqDataSet from count matrix and labels
dds <- DESeqDataSetFromMatrix(countData = counts22sub,
                              colData = colData, design = ~ condition)


# Run the default analysis for DESeq2 and generate results table
dds <- DESeq(dds)
deseq2.res <- results(dds)

# Sort by adjusted p-value and display
resOrdered <- deseq2.res[order(deseq2.res$padj), ]
vsd <- vst(dds, blind = FALSE)
exp22PCA <- plotPCA(vsd, intgroup = "condition")
nudge <- position_nudge(y = 4)
exp22PCA + geom_text(aes(label = name), position = nudge)
```

```{r}
# Select top 50 differentially expressed genes
res <- results(dds)
res_ordered <- res[order(res$padj), ]
top_genes <- row.names(res_ordered)[1:50]

# Extract counts and normalize
counts <- counts(dds, normalized = TRUE)
counts_top <- counts[top_genes, ]

# Log-transform counts
log_counts_top <- log2(counts_top + 1)

# Generate heatmap
pheatmap(log_counts_top, scale = "row")

```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ])
```
6237 degs!
just comparing exposed and control

```{r}
exp22 <- deseq2.res
# The main plot
plot(exp22$baseMean, exp22$log2FoldChange, pch=20, cex=0.45, ylim=c(-3, 3), log="x", col="darkgray",
     main="DEG SSWD Exposed  (pval <= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
exp22.sig <- deseq2.res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ]
points(exp22.sig$baseMean, exp22.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

Write out differentially expressed gene list: 
```{r}
#write.table(exp22.sig, "../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed.tab", sep = '\t', row.names = T, quote = F)
```
comment out the code - wrote out 2024-10-16

## Compare Exposed to Control Taking Age Into Account

```{r}
colDataCvEwA <- data.frame(condition=factor(c("exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "control", "control", "control", "control", "control", "control")), 
                             type=factor(rep("paired-end", 12)),
                             age=factor(c("juvenile", "juvenile", "juvenile", "adult", "adult", "adult", "juvenile", "juvenile", "juvenile", "adult", "adult", "adult")))
rownames(colDataCvEwA) <- colnames(counts22sub)
```

```{r}
# Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colDataCvEwA) %in% colnames(counts22sub)) #should return TRUE
```

```{r}
counts22sub <- counts22sub[, rownames(colDataCvEwA)]
all(rownames(colDataCvEwA) == colnames(counts22sub)) # This should also return TRUE
```

modelled after crab paper https://github.com/RobertsLab/paper-tanner-crab/blob/master/scripts/DESeq.Rmd 

```{r}
# Create a DESeqDataSet from count matrix and labels, with the design taking age into account
ddsCvEwA <- DESeqDataSetFromMatrix(countData = counts22sub,
                                     colData = colDataCvEwA, 
                                     design = ~ condition + age)
```

Check levels of age and condition 
```{r}
levels(ddsCvEwA$age)
```


```{r}
levels(ddsCvEwA$condition)
```

note: `DESeq2` automatically puts the levels in alphabetical order and the first listed level is the reference level for the factor. 

So, Control is reference. 

The following will pull the results from `condition` because that is our variable of interest. This tells us how age contributes to the DEGs
```{r}
design(ddsCvEwA) <- formula(~ age + condition)
ddsCvEwA <- DESeq(ddsCvEwA)
```

Access results:
```{r}
ddsCvEwA.res <- results(ddsCvEwA)
head(ddsCvEwA.res)
```


```{r}
summary(ddsCvEwA.res)
```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(ddsCvEwA.res[!is.na(ddsCvEwA.res$padj) & ddsCvEwA.res$padj <= 0.05, ])
```
6202

```{r}
CvEwA <- ddsCvEwA.res
# The main plot
plot(CvEwA$baseMean, CvEwA$log2FoldChange, pch=20, cex=0.45, ylim=c(-15, 15), log="x", col="darkgray",
     #main="Infection Status  (pval </= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
CvEwA.sig <- ddsCvEwA.res[!is.na(ddsCvEwA.res$padj) & ddsCvEwA.res$padj <= 0.05, ]
points(CvEwA.sig$baseMean, CvEwA.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

Write out differentially expressed gene list: 
```{r}
#write.table(CvEwA.sig, "../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed_withAge.tab", sep = '\t', row.names = T, quote = FALSE)
```

comment out 2024-10-16

## Age Contrast:

Now need to perform a contrast to see if there is a difference between these groups as it relates to age. 

In the multifactor section of the `DESeq2` manual:                
The contrast argument of the function _results_ needs a character vector of three componenets: the name of the variable (in this case "age"), and the name of the factor level for the numerator of the log2 ratio (juvenile) and the denominator (adult) 

A **contrast** is a linear combination of estimated log2 fold changes. Can be used to test if differences between groups are equal to zero. ‹        
```{r}
resultsNames(ddsCvEwA)
```

```{r}
deseq2.resage <- results(ddsCvEwA,
                          contrast = c("age", "juvenile",  "adult"))
head(deseq2.resage)
```

```{r}
age <- deseq2.resage
# The main plot
plot(age$baseMean, age$log2FoldChange, pch=20, cex=0.45, ylim=c(-15, 15), log="x", col="darkgray",
     #main="Age  (pval </= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
age.sig <- deseq2.resage[!is.na(deseq2.resage$padj) & deseq2.resage$padj <= 0.05, ]
points(age.sig$baseMean, age.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(deseq2.resage[!is.na(deseq2.resage$padj) & deseq2.resage$padj <= 0.05, ])
```

82

```{r}
summary(deseq2.resage)
```

write out DEGs from age contrast:
```{r}
#write.table(age.sig, "../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed_ageContrast.tab", sep = '\t', row.names = T, quote = FALSE)
```

wrote out 2024-10-16

## Compare Exposed to Control Taking Size Into Account (as a proxy for age)
2025-06-03

```{r}
colDataCvEwS <- data.frame(condition=factor(c("exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "control", "control", "control", "control", "control", "control")), 
                             type=factor(rep("paired-end", 12)),
                             size=factor(c("10cm", "10cm", "10cm", "32cm", "44cm", "34cm", "12cm", "8cm", "14cm", "36cm", "38cm", "42cm")))
rownames(colDataCvEwS) <- colnames(counts22sub)
```


```{r}
# Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colDataCvEwS) %in% colnames(counts22sub)) #should return TRUE
```

```{r}
counts22sub <- counts22sub[, rownames(colDataCvEwS)]
all(rownames(colDataCvEwS) == colnames(counts22sub)) # This should also return TRUE
```

modelled after crab paper https://github.com/RobertsLab/paper-tanner-crab/blob/master/scripts/DESeq.Rmd 

```{r}
# Create a DESeqDataSet from count matrix and labels, with the design taking age into account
ddsCvEwS <- DESeqDataSetFromMatrix(countData = counts22sub,
                                     colData = colDataCvEwS, 
                                     design = ~ condition + size)
```

Check levels of age and condition 
```{r}
levels(ddsCvEwS$size)
```


```{r}
levels(ddsCvEwS$condition)
```

note: `DESeq2` automatically puts the levels in alphabetical order and the first listed level is the reference level for the factor. 

So, Control is reference. 

The following will pull the results from `condition` because that is our variable of interest. This tells us how age contributes to the DEGs
```{r}
design(ddsCvEwS) <- formula(~ size + condition)
ddsCvEwS <- DESeq(ddsCvEwS)
```

Access results:
```{r}
ddsCvEwS.res <- results(ddsCvEwS)
head(ddsCvEwS.res)
```


```{r}
summary(ddsCvEwS.res)
```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(ddsCvEwS.res[!is.na(ddsCvEwS.res$padj) & ddsCvEwS.res$padj <= 0.05, ])
```
6202

```{r}
CvEwS <- ddsCvEwS.res
# The main plot
plot(CvEwA$baseMean, CvEwS$log2FoldChange, pch=20, cex=0.45, ylim=c(-15, 15), log="x", col="darkgray",
     #main="Infection Status  (pval </= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
CvEwS.sig <- ddsCvEwS.res[!is.na(ddsCvEwS.res$padj) & ddsCvEwS.res$padj <= 0.05, ]
points(CvEwS.sig$baseMean, CvEwS.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```
