---
title: "22-deseq2-2022"
output: html_document
date: "2024-10-11"
---
Rmd to take gene count matrix from [code/20-hisat2-genecount-matrices.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/20-hisat2-genecount-matrices.Rmd) through `DESeq2` to get a list or lists of differentially expressed genes. 

This was done in a previous Rmd ([code/15-hisat2-deseq2-summer2022.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/15-hisat2-deseq2-summer2022.Rmd)), but doing in a new document because I'm using a new count matrix. 

```{r}
library(DESeq2)
library(dplyr)
library(ggplot2)
library(pheatmap)
```


| library_ID | star_ID | star_age_class | treatment | experiment_day | total_armdrop | total_armtwist | disease_sign  |
|------------|---------|----------------|-----------|----------------|---------------|----------------|---------------|
| PSC.0186   |      15 | adult          | control   |             11 |             0 |              0 | healthy       |
| PSC.0203   |      31 | adult          | control   |             12 |             0 |              0 | healthy       |
| PSC.0209   |       9 | adult          | control   |             13 |             0 |              0 | healthy       |
| PSC.0177   |      34 | juvenile       | control   |             11 |             0 |              0 | healthy       |
| PSC.0219   |      63 | juvenile       | control   |             14 |             0 |              0 | healthy       |
| PSC.0230   |      65 | juvenile       | control   |             15 |             0 |              0 | healthy       |
| PSC.0174   |      30 | adult          | exposed   |             10 |             1 |              6 | first_armdrop |
| PSC.0190   |      10 | adult          | exposed   |             11 |             1 |              0 | first_armdrop |
| PSC.0231   |      13 | adult          | exposed   |             15 |             1 |              5 | first_armdrop |
| PSC.0187   |      57 | juvenile       | exposed   |             11 |             1 |              1 | first_armdrop |
| PSC.0188   |      38 | juvenile       | exposed   |             11 |             1 |              0 | first_armdrop |
| PSC.0228   |      61 | juvenile       | exposed   |             14 |             1 |              0 | first_armdrop |

Read in gene count matrix:
```{r}
counts22 <- read.csv("../data/gene_count_matrix_2022.csv")
head(counts22)
```

subset just the counts for the libraries specified in the table above:
```{r}
counts22sub <- select(counts22, gene_id, PSC.0228, PSC.0187, PSC.0188, PSC.0174, PSC.0190, PSC.0231, PSC.0230, PSC.0219, PSC.0177, PSC.0186, PSC.0209, PSC.0203)
head(counts22sub)
```
23464 rows, 13 columns (12 libraries) 

set column one (gene_id) as rownames: 
```{r}
counts22sub <- counts22sub %>% remove_rownames %>% column_to_rownames(var='gene_id')
head(counts22sub)
```
## Compare Exposed to Control 

Start out by just comparing exposed and control without doing anything with age. 
```{r}
colData <- data.frame(condition=factor(c("exposed", "exposed", "exposed", "exposed", "exposed", "exposed", "control", "control", "control", "control", "control", "control")),
                      type=factor(rep("paired-end",12)))
rownames(colData) <- colnames(counts22sub)
dds <- DESeqDataSetFromMatrix(countData = counts22sub,
                              colData = colData,
                              design = ~ condition)
```


```{r}
# Check all sample IDs in colData are also in CountData and match their orders
all(rownames(colData) %in% colnames(counts22sub)) #should return TRUE
```

```{r}
counts22sub <- counts22sub[, rownames(colData)]
all(rownames(colData) == colnames(counts22sub)) # This should also return TRUE
```

```{r}
# Create a DESeqDataSet from count matrix and labels
dds <- DESeqDataSetFromMatrix(countData = counts22sub,
                              colData = colData, design = ~ condition)


# Run the default analysis for DESeq2 and generate results table
dds <- DESeq(dds)
deseq2.res <- results(dds)

# Sort by adjusted p-value and display
resOrdered <- deseq2.res[order(deseq2.res$padj), ]
vsd <- vst(dds, blind = FALSE)
exp22PCA <- plotPCA(vsd, intgroup = "condition")
nudge <- position_nudge(y = 4)
exp22PCA + geom_text(aes(label = name), position = nudge)
```

```{r}
# Select top 50 differentially expressed genes
res <- results(dds)
res_ordered <- res[order(res$padj), ]
top_genes <- row.names(res_ordered)[1:50]

# Extract counts and normalize
counts <- counts(dds, normalized = TRUE)
counts_top <- counts[top_genes, ]

# Log-transform counts
log_counts_top <- log2(counts_top + 1)

# Generate heatmap
pheatmap(log_counts_top, scale = "row")

```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ])
```
6237 degs!
just comparing exposed and control

```{r}
exp22 <- deseq2.res
# The main plot
plot(exp22$baseMean, exp22$log2FoldChange, pch=20, cex=0.45, ylim=c(-3, 3), log="x", col="darkgray",
     main="DEG SSWD Exposed  (pval <= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
exp22.sig <- deseq2.res[!is.na(deseq2.res$padj) & deseq2.res$padj <= 0.05, ]
points(exp22.sig$baseMean, exp22.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

## Compare Exposed to Control Taking Age (Juvenile vs Adult) Into Account
Modeled after work done in my MS --> 











