---
title: "23-annotating-deg-lists"
output: html_document
date: "2024-10-14"
---
Rmd to annotated DEG lists that from [21-deseq2-2021.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/21-deseq2-2021.Rmd) and [22-deseq2-2022.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/22-deseq2-2022.Rmd). 

Packages needed:
```{r}
library(dplyr)
library(tidyverse)
```

# Get Annotation files into R environment to `join` with DEG lists: 
Files needed:

- BLAST output 
- uniprot
- count matrix 

Read in the BLAST output:  
Code where BLAST was done: [paper-pycno-sswd-2021-2022/code/16-blast-genome-annotation.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/16-blast-genome-annotation.Rmd)
```{r}
blast <- read.table("../analyses/16-blast-annotation/blast_out_sep_genes.tab", )
head(blast)
```
Rename columns: 
```{r}
colnames(blast) <- c("gene_id", "V2", "uniprot_accession", "gene_name", "V5", "V6", "V7", "V8", "V9", "V10", "V11", "V12", "V13", "V14")
head(blast)
```


Downloaded from uniprot.org ID mapping tool (copy and pasted the uniprot accession IDs from the BLAST output into the ID Mapping tool, downloaded as .tsv, then pasted into 23-annotating-deg-lists analyses folder).
```{r}
unigo <- read.delim("../analyses/23-annotating-deg-lists/idmapping_2024_10_15.tsv")
head(unigo)
```
Rename column 1 to match BLAST output file: 
```{r}
colnames(unigo) <- c("uniprot_accession", "Entry", "Reviewed", "Entry.Name", "Protein.names", "Gene.Names", "Organism", "Length", "Gene.Ontology.IDs")
head(unigo)
```

2021 Count Matrix: 

```{r}
counts21 <- read.csv("../data/gene_count_matrix_2021.csv")
head(counts21)
```
Subset the counts for the ones that the DEG lists was made from comparing: 
```{r}
exp <- select(counts21, "gene_id", "PSC.56", "PSC.52", "PSC.54", "PSC.61", "PSC.64", "PSC.73", "PSC.76", "PSC.81", "PSC.59", "PSC.57", "PSC.69", "PSC.67", "PSC.71", "PSC.75", "PSC.78", "PSC.83")
head(exp)
```

# 2021 DEG list annotation
Read in the DEG list: 
```{r}
DEG <- read.table("../analyses/21-deseq2-2021/DEGlist_2021_exposedVcontrol.tab")
head(DEG)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
DEG <- tibble::rownames_to_column(DEG, "gene_id")
head(DEG)
```
`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
DEGcount <- left_join(DEG, exp, by = "gene_id")
head(DEGcount)
```
6938 rows (GOOD) and 23 columns. 

`join` with the BLAST output based on "gene_id" column: 
```{r}
DEGcountblast <- left_join(DEGcount, blast, by = "gene_id")
head(DEGcountblast)
```
6398 rows (YAY), and 36 columns. 

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
DEGcountblastGO <- left_join(DEGcountblast, unigo, by = "uniprot_accession")
head(DEGcountblastGO)
```
6398 rows (PHEW) and 44 columns. 

WRite the table out!
```{r}
#write.table(DEGcountblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2021_exposedVcontrol_annotated.tab", sep = '\t', row.names = T, quote = F)
```
Wrote out 2024-10-15. Comment out code. 





