---
title: "23-annotating-deg-lists"
output: html_document
date: "2024-10-14"
---
Rmd to annotated DEG lists that from [21-deseq2-2021.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/21-deseq2-2021.Rmd) and [22-deseq2-2022.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/22-deseq2-2022.Rmd). 

Packages needed:
```{r}
library(dplyr)
library(tidyverse)
```

# Get Annotation files into R environment to `join` with DEG lists: 
Files needed:

- BLAST output 
- uniprot
- count matrix 

Read in the BLAST output:  
Code where BLAST was done: [paper-pycno-sswd-2021-2022/code/16-blast-genome-annotation.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/16-blast-genome-annotation.Rmd)
```{r}
blast <- read.table("../analyses/16-blast-annotation/blast_out_sep_genes.tab", )
head(blast)
```
Rename columns: 
```{r}
colnames(blast) <- c("gene_id", "V2", "uniprot_accession", "gene_name", "V5", "V6", "V7", "V8", "V9", "V10", "V11", "V12", "V13", "V14")
head(blast)
```


Downloaded from uniprot.org ID mapping tool (copy and pasted the uniprot accession IDs from the BLAST output into the ID Mapping tool, downloaded as .tsv, then pasted into 23-annotating-deg-lists analyses folder).
```{r}
unigo <- read.delim("../analyses/23-annotating-deg-lists/idmapping_2024_10_25_GOBPterms.tsv")
head(unigo)
```
Rename column 1 to match BLAST output file: 
```{r}
colnames(unigo) <- c("uniprot_accession", "Entry", "Reviewed", "Entry.Name", "Protein.names", "Gene.Names", "Organism", "Length", "Gene.Ontology.biological.process", "Gene.Ontology.IDs")
head(unigo)
```

2021 Count Matrix: 

```{r}
counts21 <- read.csv("../data/gene_count_matrix_2021.csv")
head(counts21)
```
Subset the counts for the ones that the DEG lists was made from comparing: 
```{r}
exp <- select(counts21, "gene_id", "PSC.56", "PSC.52", "PSC.54", "PSC.61", "PSC.64", "PSC.73", "PSC.76", "PSC.81", "PSC.59", "PSC.57", "PSC.69", "PSC.67", "PSC.71", "PSC.75", "PSC.78", "PSC.83")
head(exp)
```
2022 count matrix:
```{r}
counts22 <- read.csv("../data/gene_count_matrix_2022.csv")
head(counts22)
```
subset just the counts for the libraries specified in the table above:
```{r}
counts22sub <- select(counts22, gene_id, PSC.0228, PSC.0187, PSC.0188, PSC.0174, PSC.0190, PSC.0231, PSC.0230, PSC.0219, PSC.0177, PSC.0186, PSC.0209, PSC.0203)
head(counts22sub)
```



# 2021 DEG list annotation
Read in the DEG list: 
```{r}
DEG <- read.table("../analyses/21-deseq2-2021/DEGlist_2021_exposedVcontrol.tab")
head(DEG)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
DEG <- tibble::rownames_to_column(DEG, "gene_id")
head(DEG)
```
`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
DEGcount <- left_join(DEG, exp, by = "gene_id")
head(DEGcount)
```
6938 rows (GOOD) and 23 columns. 

`join` with the BLAST output based on "gene_id" column: 
```{r}
DEGcountblast <- left_join(DEGcount, blast, by = "gene_id")
head(DEGcountblast)
```
6398 rows (YAY), and 36 columns. 

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
DEGcountblastGO <- left_join(DEGcountblast, unigo, by = "uniprot_accession")
head(DEGcountblastGO)
```
6398 rows (PHEW) and 44 columns. 

WRite the table out!
```{r}
#write.table(DEGcountblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2021_exposedVcontrol_annotated.tab", sep = '\t', row.names = F, quote = F)
```
Wrote out 2024-10-15. Comment out code. 


# 2022 DEG Lists

## Control 6 vs Exposed 6
```{r}
deg1 <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed.tab")
head(deg1)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
deg1 <- tibble::rownames_to_column(deg1, "gene_id")
head(deg1)
```

`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deg1count <- left_join(deg1, counts22sub, by = "gene_id")
head(deg1count)
```

`join` with the BLAST output based on "gene_id" column: 
```{r}
deg1countblast <- left_join(deg1count, blast, by = "gene_id")
head(deg1countblast)
```
`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deg1countblastGO <- left_join(deg1countblast, unigo, by = "uniprot_accession")
head(deg1countblastGO)
```
WRite the table out!
```{r}
#write.table(deg1countblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_annotated.tab", sep = '\t', row.names = F, quote = F)
```
Wrote out 2024-10-18. Comment out code. 


## 2022 Control V Exposed taking Age into account
```{r}
deg2 <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed_withAge.tab")
head(deg2)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
deg2 <- tibble::rownames_to_column(deg2, "gene_id")
head(deg2)
```
`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deg2count <- left_join(deg2, counts22sub, by = "gene_id")
head(deg2count)
```

`join` with the BLAST output based on "gene_id" column: 
```{r}
deg2countblast <- left_join(deg2count, blast, by = "gene_id")
head(deg2countblast)
```

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deg2countblastGO <- left_join(deg2countblast, unigo, by = "uniprot_accession")
head(deg2countblastGO)
```
WRite the table out!
```{r}
#write.table(deg2countblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_withAge_annotated.tab", sep = '\t', row.names = F, quote = F)
```
Wrote out 2024-10-18. Comment out code. 


## Age contrast
```{r}
deg3 <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed_ageContrast.tab")
head(deg3)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
deg3 <- tibble::rownames_to_column(deg3, "gene_id")
head(deg3)
```


`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deg3count <- left_join(deg3, counts22sub, by = "gene_id")
head(deg3count)
```


`join` with the BLAST output based on "gene_id" column: 
```{r}
deg3countblast <- left_join(deg3count, blast, by = "gene_id")
head(deg3countblast)
```

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deg3countblastGO <- left_join(deg3countblast, unigo, by = "uniprot_accession")
head(deg3countblastGO)
```


WRite the table out!
```{r}
#write.table(deg3countblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_withAgeContrast_annotated.tab", sep = '\t', row.names = F, quote = F)
```
Wrote out 2024-10-18. Comment out code. 

## Interaction of Age with Exposure (from 2025-09-10)
```{r}
deglist <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed-with-age-interaction.tab")
head(deglist)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
deglist <- tibble::rownames_to_column(deglist, "gene_id")
head(deglist)
```


`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deglistcount <- left_join(deglist, counts22sub, by = "gene_id")
head(deglistcount)
```


`join` with the BLAST output based on "gene_id" column: 
```{r}
deglistcountblast <- left_join(deglistcount, blast, by = "gene_id")
head(deglistcountblast)
```

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deglistcountblastGO <- left_join(deglistcountblast, unigo, by = "uniprot_accession")
head(deglistcountblastGO)
```


WRite the table out!
```{r}
#write.table(deglistcountblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_withAge-interaction_annotated.tab", sep = '\t', row.names = F, quote = F)
```

# Sept 15, 2025
Annotate 8 DEGs that were found from the days of arm twist and age of 2022 experiment 

subset just the counts for the libraries specified in the table above, and JUST the exposed:
```{r}
counts22expose <- select(counts22, gene_id, PSC.0228, PSC.0187, PSC.0188, PSC.0174, PSC.0190, PSC.0231)
head(counts22expose)
```

read in 8 DEGs table:
```{r}
deglisttwist <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_armtwistduration_age.tab")
head(deglisttwist)
```

make rownames a column called "gene_id":
```{r}
library(tibble)
deglisttwist <- tibble::rownames_to_column(deglisttwist, "gene_id")
head(deglisttwist)
```

`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deglistcounttwist <- left_join(deglisttwist, counts22expose, by = "gene_id")
head(deglistcounttwist)
```

`join` with the BLAST output based on "gene_id" column: 
```{r}
deglistcountblasttwist <- left_join(deglistcounttwist, blast, by = "gene_id")
head(deglistcountblasttwist)
```

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deglistcountblastGOtwist <- left_join(deglistcountblasttwist, unigo, by = "uniprot_accession")
head(deglistcountblastGOtwist)
```

WRite the table out!
```{r}
#write.table(deglistcountblastGOtwist, "../analyses/23-annotating-deg-lists/DEGlist_2022_armtwistduration_age_annotated.tab", sep = '\t', row.names = F, quote = F)
```


# Annotate 140DEGs (age interaction) with GO Slim

read in the 140 DEG list annotated with BLAST
```{r}
agin <- read.delim("../analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_withAge-interaction_annotated.tab")
head(agin)
```
### Following this from the RobertsLab Handbook: 
[https://robertslab.github.io/resources/bio-Annotation/#map-go-ids-to-goslims](https://robertslab.github.io/resources/bio-Annotation/#map-go-ids-to-goslims)   

```{r setup, include=TRUE}
library(GSEABase)
library(GO.db)
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```


Variables

IMPORTANT: The user needs to provide:

names of the columns containing the GO IDs and the gene IDs!

Path or URL to input file.

After that, there's almost no need to modify any of the chunks which follow.

```{r set-variables, eval=TRUE}
# Column names corresponding to gene name/ID and GO IDs
GO.ID.column <- "Gene.Ontology.IDs"
gene.ID.column <- "gene_id"

# Relative path or URL to input file
input.file <- "https://raw.githubusercontent.com/grace-ac/paper-pycno-sswd-2021-2022/refs/heads/main/analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_withAge-interaction_annotated.tab"


##### Official GO info - no need to change #####
goslims_obo <- "goslim_generic.obo"
goslims_url <- "http://current.geneontology.org/ontology/subsets/goslim_generic.obo"
```


Set GSEAbase location and download goslim_generic.obo
```{r download-generic-goslim-obo, eval=TRUE}
# Find GSEAbase installation location
gseabase_location <- find.package("GSEABase")

# Load path to GOslim OBO file
goslim_obo_destintation <- file.path(gseabase_location, "extdata", goslims_obo, fsep = "/")

# Download the GOslim OBO file
download.file(url = goslims_url, destfile = goslim_obo_destintation)

# Loads package files
gseabase_files <- system.file("extdata", goslims_obo, package="GSEABase")
```

Read in gene/GO file
```{r read-in-gene-file, eval=TRUE}
full.gene.df <- read.csv(file = input.file, header = TRUE, sep = "\t")

str(full.gene.df)
```

Remove rows with NA, remove whitespace in GO IDs column and keep just gene/GO IDs columns
```{r remove-NA-and-uniprotIDs, eval=TRUE}

# Clean whitespace, filter NA/empty rows, select columns, and split GO terms using column name variables
gene.GO.df <- full.gene.df %>%
  mutate(!!GO.ID.column := str_replace_all(.data[[GO.ID.column]], "\\s*;\\s*", ";")) %>% # Clean up spaces around ";"
  filter(!is.na(.data[[gene.ID.column]]) & !is.na(.data[[GO.ID.column]]) & .data[[GO.ID.column]] != "") %>% 
  dplyr::select(all_of(c(gene.ID.column, GO.ID.column)))


str(gene.GO.df)
```

This flattens the file so all of the GO IDs per gene are separated into one GO ID per gene per row.
```{r flatten-gene-and-GO-IDs, eval=TRUE}
flat.gene.GO.df <- gene.GO.df %>% separate_rows(!!sym(GO.ID.column), sep = ";")

str(flat.gene.GO.df)
```

Groups the genes by GO ID (i.e. lists all genes associated with each unique GO ID)\
```{r group-by-GO, eval=TRUE}
grouped.gene.GO.df <- flat.gene.GO.df %>%
  group_by(!!sym(GO.ID.column)) %>%
  summarise(!!gene.ID.column := paste(.data[[gene.ID.column]], collapse = ","))

str(grouped.gene.GO.df)
```

Map GO IDs to GOslims

The mapping steps were derived from this bioconductor forum response
```{r vectorize-GOIDs, eval=TRUE}
# Vector of GO IDs
go_ids <- grouped.gene.GO.df[[GO.ID.column]]

str(go_ids)
```

Creates new OBO Collection object of just GOslims, based on provided GO IDs.
```{r extract-GOslims-from-OBO, eval=TRUE}

# Create GSEAbase GOCollection using `go_ids`
myCollection <- GOCollection(go_ids)

# Retrieve GOslims from GO OBO file set
slim <- getOBOCollection(gseabase_files)

str(slim)
```

Get Biological Process (BP) GOslims associated with provided GO IDs.
```{r retrieve-BP-GOslims, eval=TRUE}
# Retrieve Biological Process (BP) GOslims
slimdf <- goSlim(myCollection, slim, "BP", verbose)
str(slimdf)
```

Performs mapping of of GOIDs to GOslims

Returns:

GOslim IDs (as rownames)
GOslim terms
Counts of GO IDs matching to corresponding GOslim
Percentage of GO IDs matching to corresponding GOslim
GOIDs mapped to corresponding GOslim, in a semi-colon delimited format

```{r map-GO-to-GOslims, eval=TRUE}
# List of GOslims and all GO IDs from `go_ids`
gomap <- as.list(GOBPOFFSPRING[rownames(slimdf)])

# Maps `go_ids` to matching GOslims
mapped <- lapply(gomap, intersect, ids(myCollection))

# Append all mapped GO IDs to `slimdf`
# `sapply` needed to apply paste() to create semi-colon delimited values
slimdf$GO.IDs <- sapply(lapply(gomap, intersect, ids(myCollection)), paste, collapse=";")

# Remove "character(0) string from "GO.IDs" column
slimdf$GO.IDs[slimdf$GO.IDs == "character(0)"] <- ""

# Add self-matching GOIDs to "GO.IDs" column, if not present
for (go_id in go_ids) {
  # Check if the go_id is present in the row names
  if (go_id %in% rownames(slimdf)) {
    # Check if the go_id is not present in the GO.IDs column
    # Also removes white space "trimws()" and converts all to upper case to handle
    # any weird, "invisible" formatting issues.
    if (!go_id %in% trimws(toupper(strsplit(slimdf[go_id, "GO.IDs"], ";")[[1]]))) {
      # Append the go_id to the GO.IDs column with a semi-colon separator
      if (length(slimdf$GO.IDs) > 0 && nchar(slimdf$GO.IDs[nrow(slimdf)]) > 0) {
        slimdf[go_id, "GO.IDs"] <- paste0(slimdf[go_id, "GO.IDs"], "; ", go_id)
      } else {
        slimdf[go_id, "GO.IDs"] <- go_id
      }
    }
  }
}

str(slimdf)
```

"Flatten" file so each row is single GO ID with corresponding GOslim rownames_to_column needed to retain row name info
```{r flatten-GOslims-file, eval=TRUE}
# "Flatten" file so each row is single GO ID with corresponding GOslim
# rownames_to_column needed to retain row name info
slimdf_separated <- as.data.frame(slimdf %>%
  rownames_to_column('GOslim') %>%
  separate_rows(GO.IDs, sep = ";"))

# Group by unique GO ID
grouped_slimdf <- slimdf_separated %>%
  filter(!is.na(GO.IDs) & GO.IDs != "") %>%
  group_by(GO.IDs) %>%
  summarize(GOslim = paste(GOslim, collapse = ";"),
            Term = paste(Term, collapse = ";"))


str(grouped_slimdf)
```

Sorts GOslims by Count, in descending order and then selects just the Term and Count columns.
```{r sort-and-select-slimdf-counts, eval=TRUE}

slimdf.sorted <- slimdf %>% arrange(desc(Count))

slim.count.df <- slimdf.sorted %>% 
  dplyr::select(Term, Count)

str(slim.count.df)
```


write out the "slim.count.df" as a table 
```{r}
#write.table(slim.count.df, "../analyses/23-annotating-deg-lists/GOSlim-ageinteraction.tab", sep = '\t', row.names = F, quote = F)
```
wrote out 2026-03-11





