---
title: "23-annotating-deg-lists"
output: html_document
date: "2024-10-14"
---
Rmd to annotated DEG lists that from [21-deseq2-2021.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/21-deseq2-2021.Rmd) and [22-deseq2-2022.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/22-deseq2-2022.Rmd). 

Packages needed:
```{r}
library(dplyr)
library(tidyverse)
```

# Get Annotation files into R environment to `join` with DEG lists: 
Files needed:

- BLAST output 
- uniprot
- count matrix 

Read in the BLAST output:  
Code where BLAST was done: [paper-pycno-sswd-2021-2022/code/16-blast-genome-annotation.Rmd](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/16-blast-genome-annotation.Rmd)
```{r}
blast <- read.table("../analyses/16-blast-annotation/blast_out_sep_genes.tab", )
head(blast)
```
Rename columns: 
```{r}
colnames(blast) <- c("gene_id", "V2", "uniprot_accession", "gene_name", "V5", "V6", "V7", "V8", "V9", "V10", "V11", "V12", "V13", "V14")
head(blast)
```


Downloaded from uniprot.org ID mapping tool (copy and pasted the uniprot accession IDs from the BLAST output into the ID Mapping tool, downloaded as .tsv, then pasted into 23-annotating-deg-lists analyses folder).
```{r}
unigo <- read.delim("../analyses/23-annotating-deg-lists/idmapping_2024_10_25_GOBPterms.tsv")
head(unigo)
```
Rename column 1 to match BLAST output file: 
```{r}
colnames(unigo) <- c("uniprot_accession", "Entry", "Reviewed", "Entry.Name", "Protein.names", "Gene.Names", "Organism", "Length", "Gene.Ontology.biological.process", "Gene.Ontology.IDs")
head(unigo)
```

2021 Count Matrix: 

```{r}
counts21 <- read.csv("../data/gene_count_matrix_2021.csv")
head(counts21)
```
Subset the counts for the ones that the DEG lists was made from comparing: 
```{r}
exp <- select(counts21, "gene_id", "PSC.56", "PSC.52", "PSC.54", "PSC.61", "PSC.64", "PSC.73", "PSC.76", "PSC.81", "PSC.59", "PSC.57", "PSC.69", "PSC.67", "PSC.71", "PSC.75", "PSC.78", "PSC.83")
head(exp)
```
2022 count matrix:
```{r}
counts22 <- read.csv("../data/gene_count_matrix_2022.csv")
head(counts22)
```
subset just the counts for the libraries specified in the table above:
```{r}
counts22sub <- select(counts22, gene_id, PSC.0228, PSC.0187, PSC.0188, PSC.0174, PSC.0190, PSC.0231, PSC.0230, PSC.0219, PSC.0177, PSC.0186, PSC.0209, PSC.0203)
head(counts22sub)
```



# 2021 DEG list annotation
Read in the DEG list: 
```{r}
DEG <- read.table("../analyses/21-deseq2-2021/DEGlist_2021_exposedVcontrol.tab")
head(DEG)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
DEG <- tibble::rownames_to_column(DEG, "gene_id")
head(DEG)
```
`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
DEGcount <- left_join(DEG, exp, by = "gene_id")
head(DEGcount)
```
6938 rows (GOOD) and 23 columns. 

`join` with the BLAST output based on "gene_id" column: 
```{r}
DEGcountblast <- left_join(DEGcount, blast, by = "gene_id")
head(DEGcountblast)
```
6398 rows (YAY), and 36 columns. 

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
DEGcountblastGO <- left_join(DEGcountblast, unigo, by = "uniprot_accession")
head(DEGcountblastGO)
```
6398 rows (PHEW) and 44 columns. 

WRite the table out!
```{r}
#write.table(DEGcountblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2021_exposedVcontrol_annotated.tab", sep = '\t', row.names = F, quote = F)
```
Wrote out 2024-10-15. Comment out code. 


# 2022 DEG Lists

## Control 6 vs Exposed 6
```{r}
deg1 <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed.tab")
head(deg1)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
deg1 <- tibble::rownames_to_column(deg1, "gene_id")
head(deg1)
```

`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deg1count <- left_join(deg1, counts22sub, by = "gene_id")
head(deg1count)
```

`join` with the BLAST output based on "gene_id" column: 
```{r}
deg1countblast <- left_join(deg1count, blast, by = "gene_id")
head(deg1countblast)
```
`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deg1countblastGO <- left_join(deg1countblast, unigo, by = "uniprot_accession")
head(deg1countblastGO)
```
WRite the table out!
```{r}
#write.table(deg1countblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_annotated.tab", sep = '\t', row.names = F, quote = F)
```
Wrote out 2024-10-18. Comment out code. 


## 2022 Control V Exposed taking Age into account
```{r}
deg2 <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed_withAge.tab")
head(deg2)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
deg2 <- tibble::rownames_to_column(deg2, "gene_id")
head(deg2)
```
`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deg2count <- left_join(deg2, counts22sub, by = "gene_id")
head(deg2count)
```

`join` with the BLAST output based on "gene_id" column: 
```{r}
deg2countblast <- left_join(deg2count, blast, by = "gene_id")
head(deg2countblast)
```

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deg2countblastGO <- left_join(deg2countblast, unigo, by = "uniprot_accession")
head(deg2countblastGO)
```
WRite the table out!
```{r}
#write.table(deg2countblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_withAge_annotated.tab", sep = '\t', row.names = F, quote = F)
```
Wrote out 2024-10-18. Comment out code. 


## Age contrast
```{r}
deg3 <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed_ageContrast.tab")
head(deg3)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
deg3 <- tibble::rownames_to_column(deg3, "gene_id")
head(deg3)
```


`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deg3count <- left_join(deg3, counts22sub, by = "gene_id")
head(deg3count)
```


`join` with the BLAST output based on "gene_id" column: 
```{r}
deg3countblast <- left_join(deg3count, blast, by = "gene_id")
head(deg3countblast)
```

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deg3countblastGO <- left_join(deg3countblast, unigo, by = "uniprot_accession")
head(deg3countblastGO)
```


WRite the table out!
```{r}
#write.table(deg3countblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_withAgeContrast_annotated.tab", sep = '\t', row.names = F, quote = F)
```
Wrote out 2024-10-18. Comment out code. 

## Interaction of Age with Exposure (from 2025-09-10)
```{r}
deglist <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_controlVexposed-with-age-interaction.tab")
head(deglist)
```
make rownames a column called "gene_id":
```{r}
library(tibble)
deglist <- tibble::rownames_to_column(deglist, "gene_id")
head(deglist)
```


`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deglistcount <- left_join(deglist, counts22sub, by = "gene_id")
head(deglistcount)
```


`join` with the BLAST output based on "gene_id" column: 
```{r}
deglistcountblast <- left_join(deglistcount, blast, by = "gene_id")
head(deglistcountblast)
```

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deglistcountblastGO <- left_join(deglistcountblast, unigo, by = "uniprot_accession")
head(deglistcountblastGO)
```


WRite the table out!
```{r}
#write.table(deglistcountblastGO, "../analyses/23-annotating-deg-lists/DEGlist_2022_exposedVcontrol_withAge-interaction_annotated.tab", sep = '\t', row.names = F, quote = F)
```

# Sept 15, 2025
Annotate 8 DEGs that were found from the days of arm twist and age of 2022 experiment 

subset just the counts for the libraries specified in the table above, and JUST the exposed:
```{r}
counts22expose <- select(counts22, gene_id, PSC.0228, PSC.0187, PSC.0188, PSC.0174, PSC.0190, PSC.0231)
head(counts22expose)
```

read in 8 DEGs table:
```{r}
deglisttwist <- read.table("../analyses/22-deseq2-2022/DEGlist_2022_armtwistduration_age.tab")
head(deglisttwist)
```

make rownames a column called "gene_id":
```{r}
library(tibble)
deglisttwist <- tibble::rownames_to_column(deglisttwist, "gene_id")
head(deglisttwist)
```

`join` the DEG list with the count matrix based on the "gene_id" column: 
```{r}
deglistcounttwist <- left_join(deglisttwist, counts22expose, by = "gene_id")
head(deglistcounttwist)
```

`join` with the BLAST output based on "gene_id" column: 
```{r}
deglistcountblasttwist <- left_join(deglistcounttwist, blast, by = "gene_id")
head(deglistcountblasttwist)
```

`join` with uniprot GO list based on uniprot_accession column: 
```{r}
deglistcountblastGOtwist <- left_join(deglistcountblasttwist, unigo, by = "uniprot_accession")
head(deglistcountblastGOtwist)
```

WRite the table out!
```{r}
#write.table(deglistcountblastGOtwist, "../analyses/23-annotating-deg-lists/DEGlist_2022_armtwistduration_age_annotated.tab", sep = '\t', row.names = F, quote = F)
```
