---
title: "26-visualizations"
output: html_document
date: "2024-11-18"
---
Rmd to make visualizations for paper. 

https://yulab-smu.top/biomedical-knowledge-mining-book/enrichplot.html 

```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(RColorBrewer)
```


# Make a figure of the enriched GO Processes

Read in the list of 9 top enriched processes:
```{r}
top9 <- read.delim("../analyses/25-compare-2021-2022/2021-2022-entichedtop9_DAVID_separated.txt")
head(top9)
```
Goal: plots of the terms, x-axis is counts and color is benjamini value. 


```{r}

# Basic barplot
p <- ggplot(data=top9, aes(x=Count, y=Term, fill=Fold.Enrichment)) +
  geom_bar(stat="identity")+theme_minimal()
p

# Rename y-axis:


# Horizontal bar plot
p + coord_flip()
```

# Make a Venn Diagram of the DEGs for 2021, 2022, and the overlap... is this just numbers??

Total DEGS for 2021 --> 6938
Total amount unique to 2021 --> (6938-4114 =) 2824
Total DEGs for 2022 --> 6237
Total amount unique to 2022 --> (6237-4114 =) 2123
```{r}
#install.packages("ggVennDiagram")
library(ggVennDiagram)
```
```{r}
#install.packages("VennDiagram")
library(VennDiagram)
```



```{r}
# create pairwise Venn diagram 
draw.pairwise.venn(area1=6938, area2=6237,cross.area=4114, 
                   category=c("A","B"),fill=c("Red","Yellow"))
# 2D Venn diagram
ggVennDiagram(x) 
```

# Disease Progression and Mortality figure: 
Focusing on just the observations for the n=16 stars in 2021 (Exp A) and n=12 stars in 2022 (Exp B) 

https://cran.r-project.org/web/packages/vistime/vignettes/gg_vistime-vignette.html \    
https://shosaco.github.io/vistime/ 

 Goal -- timeline color-coded horizontal stacked bar... 

