---
title: "28-circos"
output: html_document
date: "2025-05-30"
---
Circos plot of DEGs from the study. 

🔧 Concept:
A circular ideogram with three arcs:
• Arc A: DEGs from Adult-only experiment
• Arc B: DEGs from Age-structured experiment
• Arc C: Shared DEGs, with immune-related genes highlighted
• Chords (connecting bands) link genes that appear in:
• Multiple experiments
• Multiple GO/KEGG immune pathways

It seems like Circos isn't an R package... but there is an R package called `circlize` that i'll use. 

example: https://bioinfo4all.wordpress.com/2021/03/13/tutorial-7-how-to-do-chord-diagram-using-r/ 

https://jokergoo.github.io/circlize_book/book/the-chorddiagram-function.html 

install `circlize` package:
```{r}
#install.packages("circlize")
```
installed 2025-06-02

load package and others needed:
```{r}
library(circlize)
library(tidyverse)
library(dplyr)
library(RColorBrewer)
```

View annotated DEG list from Adult-only experiment:
```{r}
adultDEGsannot <- read.delim("../supplemental/S02-DEGlist_2021_exposedVcontrol_annotated.tab")
head(adultDEGsannot)
```

resource: https://r-charts.com/flow/chord-diagram/ 
Format of data for `circlize` plot should be:

        exp 1         exp 2       shared
GO BP   #degs         #degs       #degs
GO BP   #degs         #degs       #degs

and so on

If a DEG has multiple GO BP terms associated with it, the data frame would look like: 
SOMETHING

I want the figure to look like one linked here: 
https://www.researchgate.net/figure/Chord-plots-displaying-relationships-between-RNAseq-gene-expression-data-and-gene_fig6_366322182 


there are  TON of DEGs in each list... so i need to think about how to shorten. Pick a metric and pick the top X number of DEGs in each experiment?? How best to do that? 

I'm going to play around a little bit to figure out how to use the package before using the bigger tables.



