---
title: "28-circos"
output: html_document
date: "2025-05-30"
---
Circos plot of DEGs from the study. 

🔧 Concept:
A circular ideogram with three arcs:
• Arc A: DEGs from Adult-only experiment
• Arc B: DEGs from Age-structured experiment
• Arc C: Shared DEGs, with immune-related genes highlighted
• Chords (connecting bands) link genes that appear in:
• Multiple experiments
• Multiple GO/KEGG immune pathways

It seems like Circos isn't an R package... but there is an R package called `circlize` that i'll use. 

example: https://bioinfo4all.wordpress.com/2021/03/13/tutorial-7-how-to-do-chord-diagram-using-r/ 

https://jokergoo.github.io/circlize_book/book/the-chorddiagram-function.html 

install `circlize` package:
```{r}
#install.packages("circlize")
```
installed 2025-06-02

load package and others needed:
```{r}
library(circlize)
library(tidyverse)
library(dplyr)
library(RColorBrewer)
```

View annotated DEG list from Adult-only experiment:
```{r}
adultDEGsannot <- read.delim("../supplemental/S02-DEGlist_2021_exposedVcontrol_annotated.tab")
head(adultDEGsannot)
```

resource: https://r-charts.com/flow/chord-diagram/ 
Format of data for `circlize` plot should be:

        exp 1         exp 2       shared
GO BP   #degs         #degs       #degs
GO BP   #degs         #degs       #degs

and so on

If a DEG has multiple GO BP terms associated with it, the data frame would look like: 
SOMETHING

I want the figure to look like one linked here: 
https://www.researchgate.net/figure/Chord-plots-displaying-relationships-between-RNAseq-gene-expression-data-and-gene_fig6_366322182 


there are  TON of DEGs in each list... so i need to think about how to shorten. Pick a metric and pick the top X number of DEGs in each experiment?? How best to do that? 

I'm going to play around a little bit to figure out how to use the package before using the bigger tables.

load practice data:
```{r}
data <- read.delim("../analyses/28-chord-plots/practice_data.csv", sep = ",")
head(data)
```
make first column into rownames:
```{r}
library(tidyverse)

#use values from 'X' column as row names
data <- data %>% remove_rownames %>% column_to_rownames(var='X')

```

make the data into a matrix:
```{r}
# Converting into numeric matrix
data2 <- data.matrix(data)
data2
```


Create plot:
```{r}
# install.packages("circlize")
library(circlize)

chordDiagram(data2)

```

Wait...

actually... i could have this just be of the enriched processes (9) and the DEGs associated with them!!!! maybe that would be cool? 

try (note: several DEGs in each list have multiple annotations - need to figure out how to do split chords)
```{r}
data <- read.delim("../analyses/28-chord-plots/top9-degs-for-circle-plot.csv", sep = ",")
head(data)
```
make first column into rownames:
```{r}
library(tidyverse)

#use values from 'X' column as row names
data <- data %>% remove_rownames %>% column_to_rownames(var='X')

```

make the data into a matrix:
```{r}
# Converting into numeric matrix
data2 <- data.matrix(data)
data2
```

Create plot:
```{r}
# install.packages("circlize")
library(circlize)

chordDiagram(data2)

```

Try a new orientation where it's the unique DEGs for 2021, unique DEGs for 2022, and then the shared, with chords 

# Try with unique DEGs to 2021 and 2022 separated, with shared being it's own band

```{r}
data <- read.delim("../analyses/28-chord-plots/top9-degs-for-chord-uniq20212022.csv", sep = ",")
head(data)
```

make first column into rownames:
```{r}
library(tidyverse)

#use values from 'X' column as row names
data <- data %>% remove_rownames %>% column_to_rownames(var='X')

```

make the data into a matrix:
```{r}
# Converting into numeric matrix
data2 <- data.matrix(data)
data2
```
 
 
                                                                            Adult.SSWD.Experiment Age.Class.SSWD.Experiment Shared
GO:0043161~proteasome-mediated ubiquitin-dependent protein catabolic process                    16                        15     87
GO:0002181~cytoplasmic translation                                                               6                         4     24
GO:0006508~proteolysis                                                                          48                        59    131
GO:0006511~ubiquitin-dependent protein catabolic process                                        13                        21     58
GO:0007229~integrin-mediated signaling pathway                                                   2                         4     23
GO:0033209~tumor necrosis factor-mediated signaling pathway                                      1                         3     16
GO:0042742~defense response to bacterium                                                         7                         6     28
GO:0006412~translation                                                                          14                        18     44
GO:0097352~autophagosome maturation                                                              2                         3     16


Create plot:
```{r}
# install.packages("circlize")
library(circlize)

chordDiagramFromMatrix(data2, )

```

# change of package!!! I'll now try `GOPlot` R package
https://wencke.github.io

```{r}
# Installation of the latest released version
install.packages('GOplot')
```
Installed 2025-06-02

```{r}
library(GOplot)
```
follolwing this: https://wencke.github.io 

read in the DAVID output from the top9 BP:
```{r}
davidout <- read.delim("../supplemental/S09-2021-2022-same-DAVID-top9_enriched.tab")
head(davidout)
```



load genes (from shared 4,114 of degs that are associated with the top 9 enriched terms) with log2fc
```{r}
genelog <- read.csv("../analyses/28-chord-plots/genes-log2fc.csv")
head(genelog)
```
make it a dataframe:
```{r}
genelog <- as.data.frame(genelog)
head(genelog)
```

create a vector of the top 9 enriched BP: 
```{r}
processes <- c("proteasome-mediated ubiquitin-dependent protein catabolic process", "cytoplasmic translation", "proteolysis", "ubiquitin-dependent protein catabolic process", "integrin-mediated signaling pathway", "tumor necrosis factor-mediated signaling pathway", "defense response to bacterium", "translation", "autophagosome maturation")
head(processes)
```

```{r}
# Generate the plotting object
circ <- circle_dat(davidout, genelog)
```


```{r}
# Now it is time to generate the binary matrix
chord <- chord_dat(circ, genelog, processes)
head(chord)
```



