---
title: "31-network-diagram-S11"
output: html_document
date: "2025-11-19"
---

# Network Diagram for Enriched Processes

This R Markdown document creates a network diagram showing the relationships between 
GO Biological Process terms and UniProt IDs from the supplemental file 
`S11-DEGlist_enrichedprocesses_2021-2022.tab`.

The network diagram will show:
- GO terms as one set of nodes
- UniProt IDs as another set of nodes
- Connections showing which IDs are associated with which GO terms
- This helps visualize how certain genes/proteins may be involved in multiple biological processes

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Install and Load Required Packages

```{r install-packages, eval=FALSE}
# Install packages if needed
if (!require("igraph")) install.packages("igraph")
if (!require("ggraph")) install.packages("ggraph")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("ggplot2")) install.packages("ggplot2")
```

```{r load-libraries}
library(igraph)
library(ggraph)
library(tidyverse)
library(ggplot2)
```

## Read the Data

```{r read-data}
# Read the supplemental file
data <- read.table("../supplemental/S11-DEGlist_enrichedprocesses_2021-2022.tab", 
                   header = TRUE, 
                   sep = "\t",
                   stringsAsFactors = FALSE)

# Check the structure
head(data)
dim(data)

# Extract only the first two columns
network_data <- data[, 1:2]
colnames(network_data) <- c("GO_Term", "UniProt_ID")

# Summary statistics
cat("Number of unique GO terms:", length(unique(network_data$GO_Term)), "\n")
cat("Number of unique UniProt IDs:", length(unique(network_data$UniProt_ID)), "\n")
cat("Total connections:", nrow(network_data), "\n")
```

## Prepare Network Data

```{r prepare-network}
# Create a simplified term name for better visualization
network_data$GO_Term_Short <- gsub("GO:[0-9]+~", "", network_data$GO_Term)

# Count how many times each UniProt ID appears (degree)
id_counts <- network_data %>%
  group_by(UniProt_ID) %>%
  summarise(degree = n()) %>%
  arrange(desc(degree))

cat("Top 10 most connected UniProt IDs:\n")
print(head(id_counts, 10))

# Count how many IDs are associated with each GO term
term_counts <- network_data %>%
  group_by(GO_Term_Short) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

cat("\nGO terms and their associated ID counts:\n")
print(term_counts)
```

## Create Network Graph

```{r create-graph}
# Create a bipartite graph
# We'll use igraph to create the network
edges <- network_data[, c("GO_Term_Short", "UniProt_ID")]

# Create the graph from the edge list
g <- graph_from_data_frame(edges, directed = FALSE)

# Set vertex types for bipartite layout
V(g)$type <- V(g)$name %in% network_data$GO_Term_Short

# Color nodes by type
V(g)$color <- ifelse(V(g)$type, "lightblue", "lightcoral")
V(g)$label.color <- "black"

# Size GO term nodes by number of connections
go_sizes <- term_counts$count[match(V(g)$name, term_counts$GO_Term_Short)]
id_sizes <- id_counts$degree[match(V(g)$name, id_counts$UniProt_ID)]
V(g)$size <- ifelse(V(g)$type, 
                    10 + (go_sizes / max(go_sizes, na.rm = TRUE) * 20),
                    3)

cat("Network created with", vcount(g), "nodes and", ecount(g), "edges\n")
```

## Visualize the Network - Basic Plot

```{r plot-basic, fig.width=12, fig.height=10}
# Plot the basic network
set.seed(123)  # For reproducibility
plot(g, 
     layout = layout_with_fr(g),
     vertex.label.cex = 0.7,
     vertex.frame.color = "gray50",
     edge.color = "gray80",
     main = "Network: GO Terms and UniProt IDs\n(Blue = GO Terms, Red = UniProt IDs)")
```

## Visualize with ggraph

```{r plot-ggraph, fig.width=14, fig.height=12}
set.seed(123)
# Create a more sophisticated visualization with ggraph
ggraph(g, layout = 'fr') + 
  geom_edge_link(alpha = 0.2, color = "gray50") +
  geom_node_point(aes(color = type, size = size)) +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "coral"),
                     labels = c("TRUE" = "GO Terms", "FALSE" = "UniProt IDs"),
                     name = "Node Type") +
  scale_size_continuous(range = c(1, 8), guide = "none") +
  geom_node_text(aes(label = ifelse(type, name, "")), 
                 repel = TRUE, size = 3, fontface = "bold") +
  theme_void() +
  labs(title = "Network Diagram: GO Biological Processes and Associated UniProt IDs",
       subtitle = "Lines connect UniProt IDs to their associated GO terms",
       caption = "Data: S11-DEGlist_enrichedprocesses_2021-2022.tab") +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 12, hjust = 0.5),
        legend.position = "bottom")
```

## Alternative: Bipartite Layout

```{r plot-bipartite, fig.width=14, fig.height=12}
# Create a bipartite layout showing GO terms on one side and IDs on the other
set.seed(123)
ggraph(g, layout = 'bipartite') + 
  geom_edge_link(alpha = 0.15, color = "gray60") +
  geom_node_point(aes(color = type, size = size)) +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "coral"),
                     labels = c("TRUE" = "GO Terms", "FALSE" = "UniProt IDs"),
                     name = "Node Type") +
  scale_size_continuous(range = c(1, 8), guide = "none") +
  geom_node_text(aes(label = ifelse(type, name, "")), 
                 repel = FALSE, size = 2.5, angle = 90, hjust = 0, fontface = "bold") +
  theme_void() +
  labs(title = "Bipartite Network: GO Terms ↔ UniProt IDs",
       subtitle = "GO terms on the left, UniProt IDs on the right",
       caption = "Data: S11-DEGlist_enrichedprocesses_2021-2022.tab") +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 12, hjust = 0.5),
        legend.position = "bottom")
```

## Network Statistics

```{r network-stats}
cat("Network Statistics:\n")
cat("===================\n")
cat("Nodes:", vcount(g), "\n")
cat("Edges:", ecount(g), "\n")
cat("Density:", edge_density(g), "\n")
cat("Average degree:", mean(degree(g)), "\n")
cat("\n")

# Find which GO terms share the most UniProt IDs
cat("GO Term Overlap Analysis:\n")
cat("=========================\n")

# Calculate Jaccard similarity between GO terms
go_terms <- unique(network_data$GO_Term_Short)
overlap_matrix <- matrix(0, nrow = length(go_terms), ncol = length(go_terms))
rownames(overlap_matrix) <- colnames(overlap_matrix) <- go_terms

for(i in 1:length(go_terms)) {
  for(j in i:length(go_terms)) {
    ids_i <- network_data$UniProt_ID[network_data$GO_Term_Short == go_terms[i]]
    ids_j <- network_data$UniProt_ID[network_data$GO_Term_Short == go_terms[j]]
    intersection <- length(intersect(ids_i, ids_j))
    union <- length(union(ids_i, ids_j))
    overlap_matrix[i, j] <- overlap_matrix[j, i] <- intersection / union
  }
}

# Show pairs with highest overlap (excluding diagonal)
diag(overlap_matrix) <- 0
max_overlap <- which(overlap_matrix == max(overlap_matrix), arr.ind = TRUE)[1, ]
cat("Highest overlap between:\n")
cat("  -", go_terms[max_overlap[1]], "\n")
cat("  -", go_terms[max_overlap[2]], "\n")
cat("  Jaccard similarity:", overlap_matrix[max_overlap[1], max_overlap[2]], "\n")
```

## Save the Network Diagram

```{r save-plot}
# Save the main network diagram
ggsave("../analyses/network-diagram-S11-GO-UniProt.png", 
       width = 14, height = 12, dpi = 300, bg = "white")

cat("Network diagram saved to: ../analyses/network-diagram-S11-GO-UniProt.png\n")
```

## Session Info

```{r session-info}
sessionInfo()
```
